Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-01-24 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1970 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 02f42108363b7fd61e0c06bede33de04319d1cc4

\title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}

@@

\SweaveOpts{eval = TRUE}
\section{Introduction}
In regression analysis various models have been developed to describe the relationship between a response variable and a (set of) covariate(s). Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects. In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate. In the left picture the expected value of the response variable is modeled by a linear function in the covariate. In the middle a GAM is applied where a smooth function allows for more flexibility.

@@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
LM, GLM  
\end{center}
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
GAM 
\end{center}
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
GAMLSS 
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
\end{tikzpicture}
\caption{\label{Devel_parmod}Development in the field of parametric modeling: (Generalized) linear models (LM, GLM; left), generalized additive models (GAM; middle), generalized additive models for location, scale, and shape (GAMLSS; right).}
\end{figure}

@@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar = c(2, 0, 2, 0))
plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
points(x, y, col = "slategray", pch = 19)
box(lwd=8)
@
\vspace{-0.2cm}%\\
Distributional Tree
\vspace{0.2cm}
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](7.1,0)--(7.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](0,-2.4)--(0,-3.2);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](2.2,-5.4)--(7.9,-2.5);
\end{tikzpicture}
\caption{\label{Devel_treeforest}Development in the field of algorithmic trees and forests: regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
\end{figure}

@@

# add labels
text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
@
\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008. Observations and predictions are left censored at 0. The corresponding point mass is shown at the censoring point (0).}
\end{center}
\end{figure}

@@

\subsection{Models and evaluation}
Next to distributional forests three other methods are applied to this data set and their results are compared based on the continuous ranked probability score (CRPS). One of the reference models is an EMOS model as this is a common choice in this type of setting and research field. The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameter and once applying a boosting algorithm. 
The chosen methods and their specifications are listed in Table~\ref{model_specification}.
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
%\begin{minipage}{\textwidth}
\begin{center}
\begin{tabular}{ l l l l }
\hline
Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$) \\
\hline
\textbf{distributional forest} & recursive    & all & all \\ 
                & partitioning &     &     \\ 
\hline
\textbf{prespecified GAM} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
                            & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
 & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
 & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
 & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
 & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
 & & \emph{pwat{\_}mean{\_}mean}, & \\
 & & \emph{tdiff500850{\_}mean}, & \\
 & & \emph{msl{\_}diff} & \\
\hline
\textbf{boosted GAM} & spline  & all & all \\
                       & in each &     &     \\
\hline
\textbf{EMOS} & linear & \emph{tp{\_}mean} & \emph{tp{\_}sprd} \\
\hline
\end{tabular}
\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter.}
\label{model_specification}
%}
\end{center}
\end{table}

@@

\begin{figure}[t!]
\begin{center}
%\begin{subfigure}{0.7\textwidth}
<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
par(mar = c(2,10,2,2))
barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
        horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE,
        font.axis = 3, #list(family="HersheySerif", face=3),
        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
@
%\end{subfigure}
%\hspace{-1cm}
%\begin{subfigure}{0.5\textwidth}
%<<rain_axams_varim_margeffects1_sigma, fig=TRUE, echo=FALSE>>=
%plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = %"sigma")
%@
%\end{subfigure}
\caption{\label{varimp} Variable importance of the distributional forest for station Axams learned on data from 1985--2008 and assessed in 2009--2012 based on mean decrease in accuracy regarding the CRPS (top 10 list).}
\end{center}
\end{figure}

@@

@
%\end{subfigure}
\caption{\label{pit_oosample}Out-of-sample PIT histograms for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\center
<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
        names = c("distributional forest", "prespecified GAM", "boosted GAM"),
        ylab = "CRPS skills score", col = "lightgray") 
abline(h = 0, col = pal["EMOS"], lwd = 2)
@
\center
\caption{\label{boxplots_crps_axams}CRPS skills score with reference model EMOS from the 10 times 7-fold cross validation at station Axams.}
\end{center}
\end{figure}

@@

@
\caption{\label{boxplot_crps_all}CRPS skills score over all stations with reference model EMOS. The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
\end{center}
\end{figure}

@@

@
\caption{\label{map}Map of Tyrol with all observation stations. The type of symbol indicates which model performed best with respect to the CRPS. The color indicates the difference of the CRPS of the distributional forest and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).}
\end{center}
\end{figure}

@@

# out of sample
set.seed(7)
par(mfrow = c(2,2))
set.seed(1)
qqrplot(pit_df_t, nsim = 1, main = "distributional forest", ylim = c(-5,5))
qqrplot(pit_g_t, nsim = 1, main = "prespecified GAM", ylim = c(-5,5))
qqrplot(pit_gb_t, nsim = 1, main = "boosted GAM", ylim = c(-5,5))
qqrplot(pit_ml_t, nsim = 1, main = "EMOS", ylim = c(-5,5))

@
\caption{\label{qqr_oosample} Out-of-sample QQR-plot for station Axams assessed in 2009--2012 based on models learned on data from 1985--2008.}
\end{center}
\end{figure}

>>>>

\title{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
\Shorttitle{Distributional Regression Forests for Probabilistic Precipitation Forecasting}

@@

\SweaveOpts{eval = TRUE}
\section{Introduction}
In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates. Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects. In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate. In the left picture the expected value of the response variable is modeled by a linear function in the covariate. In the middle a GAM is applied where a smooth function allows for more flexibility.

@@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage};
\node (d) at (0,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
LM, GLM  
\end{center}
\endminipage};
\node (e) at (5,-2) 
{
\minipage{0.29\textwidth}
\begin{center}
\vspace{0.0cm}
GAM 
\end{center}
\endminipage};
\node at (10,-2)
{
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.0cm}
GAMLSS 
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](7.1,0)--(7.9,0);
\end{tikzpicture}
\caption{\label{Devel_parmod}Parametric modeling developments. (Generalized) linear models (left), generalized additive models (middle), generalized additive models for location, scale, and shape (right).}
\end{figure}

@@

<<plot_motivation_disttree, fig=TRUE, echo=FALSE>>=
par(mar = c(2, 0, 2, 0))
plot(x, y, xaxt="n", yaxt="n", ann=FALSE, type = "n")
usr <- par("usr")
rect(usr[1], usr[3], usr[2], usr[4], col = gray(0.94), border = "transparent")
lines(x = x, y = ytree, col = pal["forest"], lwd=7, main = "")
polygon(c(x, rev(x)), c(ytree + var, rev(ytree - var)),
  col = pallight["forest"], border = "transparent")
points(x, y, col = "slategray", pch = 19)
box(lwd=8)
@
\vspace{-0.2cm}%\\
Distributional Tree
\vspace{0.2cm}
\end{center}
\endminipage};
\draw[-{>[scale=4, length=2, width=3]},line width=0.4pt](2.1,0)--(2.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](7.1,0)--(7.9,0);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](0,-2.4)--(0,-3.2);
\draw[-{>[scale=4, length=2, width=3]},line width=0.9pt](2.2,-5.4)--(7.9,-2.5);
\end{tikzpicture}
\caption{\label{Devel_treeforest}Tree and forest developments. Regression tree (top left), distributional tree (bottom), random forest (top middle), distributional forest (top right).}
\end{figure}

@@

# add labels
text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
@
\caption{\label{axams_onestation}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008. Observations are left censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
\end{center}
\end{figure}

@@

\subsection{Models and evaluation}
Next to distributional forests three other methods are applied to this data set and their results are compared based on the continuous ranked probability score (CRPS). One of the reference models is an EMOS model as this is a common choice in this type of setting and research field. The other two models are both GAMs, once in a prespecified form in terms of variable selection for each distribution parameter and once applying a boosting algorithm. 
The chosen methods and their specifications are listed in Table~\ref{model_specification}.
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
%\begin{minipage}{\textwidth}
\begin{center}
\begin{tabular}{ l l l l }
\hline
Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$) \\
\hline
\textbf{distributional forest} & recursive    & all & all \\ 
                & partitioning &     &     \\ 
\hline
\textbf{prespecified GAM} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
                            & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
 & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
 & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
 & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
 & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
 & & \emph{pwat{\_}mean{\_}mean}, & \\
 & & \emph{tdiff500850{\_}mean}, & \\
 & & \emph{msl{\_}diff} & \\
\hline
\textbf{boosted GAM} & spline  & all & all \\
                       & in each &     &     \\
\hline
\textbf{EMOS} & linear & \emph{tp{\_}mean} & \emph{tp{\_}sprd} \\
\hline
\end{tabular}
\caption[Table caption text]{Overview of models with type of covariate dependency and included covariates for each distribution parameter.}
\label{model_specification}
%}
\end{center}
\end{table}

@@

\begin{figure}[t!]
\begin{center}
%\begin{subfigure}{0.65\textwidth}
<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.5>>=
par(mar = c(2,10,2,2))
barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
        horiz = TRUE, las = 1, cex.names = 1.2, axes = FALSE,
        xlab = "Variable importance: mean decrease in CRPS",
        font.axis = 3, #list(family="HersheySerif", face=3),
        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
@
%\end{subfigure}
%\hspace{-1cm}
%\begin{subfigure}{0.5\textwidth}
%<<rain_axams_varim_margeffects1_sigma, fig=TRUE, echo=FALSE>>=
%plot(x = sp1[,id_mi[1]], y = sp1[, "sigma"], type = "l", xlab = names_mi[1], ylab = %"sigma")
%@
%\end{subfigure}
\caption{\label{varimp} CRPS based variable importance for the top 10 covariates in the distributional forest. Based on data for station Axams, learning period 1985--2008 and assessed in 2009--2012.}
\end{center}
\end{figure}

@@

@
%\end{subfigure}
\caption{\label{pit_oosample}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\center
<<rain_cross_axams_crps_skills_score, fig=TRUE, echo=FALSE, height=4.5, width = 6.5>>=
boxplot(1 - rain_crps[,c(2,3,4)] / rain_crps[,6], ylim = c(-0.005, 0.065),
        names = c("distributional forest", "prespecified GAM", "boosted GAM"),
        ylab = "CRPS skills score", col = "lightgray") 
abline(h = 0, col = pal["EMOS"], lwd = 2)
@
\center
\caption{\label{boxplots_crps_axams}CRPS skills score from the 10 times 7-fold cross validation at station Axams (1985--2012). The horizontal orange line pertains to the reference model EMOS.}
\end{center}
\end{figure}

@@

@
\caption{\label{boxplot_crps_all}CRPS skills score for each station (grey lines) and aggregated over all stations (boxplots). Station Axams is highlighted in greend and the horizontal orange line pertains to the reference model EMOS. The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
\end{center}
\end{figure}

@@

@
\caption{\label{map}Map of Tyrol showing which model performed best at which station (type of symbol). The color codes whether the distributional forest had lower or higher CRPS (red or blue) compared to the best of the other three models (prespecified GAM, boosted GAM and EMOS).}
\end{center}
\end{figure}

@@

# out of sample
set.seed(7)
par(mfrow = c(2,2))
qqrplot(pit_df_t, nsim = 100, main = "distributional forest", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
qqrplot(pit_g_t, nsim = 100, main = "prespecified GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
qqrplot(pit_gb_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)

qqrplot(pit_ml_t, nsim = 100, main = "boosted GAM", ylim = c(-5,5), col = gray(0.1, alpha = 0.02), pch = 19)
@
\caption{\label{qqr_oosample} Out-of-sample residual QQ-plot (2009--2012) for station Axams based on models learned on data from 1985--2008.}
\end{center}
\end{figure}
